---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, message = F, warning = F}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = F, warning = F)

library(pepreporting)
library(ggplot2)
library(mapview)
library(dplyr)
library(lubridate)
library(tidyr)
library(sf)
```


The package includes a `stations` data object that includes metadata for each station, including lat/lon and bay segment.  

```{r}
prj <- 4326
locs <- stations %>% 
  st_as_sf(coords = c('Longitude', 'Latitude'), crs = prj)

mapview(locs, zcol = 'bay_segment', layer.name = 'Bay segment')            
```

The water quality data can be imported using the `read_pepwq()` function.  A compressed folder that inludesthe data can be downloaded from [here](https://gisportal.suffolkcountyny.gov/gis/home/item.html?id=8107f192ffac406380b6d61d3d3dbf7d).  After the data are downloaded and extracted, the Excel file with the raw data is named "Peconics SCDHS WQ data - up to 2019 so far.xlsx", or something similar depending on when the data were downloaded.  The location of this file on your computer is passed to the import function. Below, a local file renamed as "currentdata.xlsx" that contains the water quality data is downloaded. 

```{r}
dat <- read_pepwq('../inst/extdata/currentdata.xlsx')
head(dat)
```

The raw data includes multiple fields, but only the chlorophyll and secchi data are retained for reporting.  Each station is grouped by major bay segment, defined as `r levels(dat$bay_segment)`. 

A quick view of the number of observations and length of record at each stations shows that effort was not continuous. 

```{r, fig.height = 7, fig.width = 11}
toplo <- dat %>% 
  select(bay_segment, BayStation, chla, sd) %>% 
  pivot_longer(c('chla', 'sd')) %>% 
  group_by(bay_segment, BayStation, name) %>% 
  summarise(`Obs. (n)` = n())

p <- ggplot(toplo, aes(x = yr, y = BayStation, fill = `Obs. (n)`)) + 
  geom_tile()+ #colour = 'lightgrey') + 
  facet_grid(bay_segment ~ name, scales = 'free_y', space = 'free_y') + 
  theme_bw() + 
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor= element_blank(),
    strip.background = element_blank(), 
    axis.title.x = element_blank(), 
    legend.position = 'top'
  ) + 
  scale_fill_viridis_c()

p
```

