---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, message = F, warning = F}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = F, warning = F)

library(pepreporting)
library(ggplot2)
library(mapview)
library(dplyr)
library(lubridate)
library(tidyr)
library(sf)
```

The package includes a `pepstations` data object that includes metadata for each station, including lat/lon and bay segment.  

```{r}
prj <- 4326
locs <- pepstations %>% 
  st_as_sf(coords = c('Longitude', 'Latitude'), crs = prj)

mapview(locs, zcol = 'bay_segment', layer.name = 'Bay segment')            
```

The water quality data can be imported using the `read_pepwq()` function.  A compressed folder that inludesthe data can be downloaded from [here](https://gisportal.suffolkcountyny.gov/gis/home/item.html?id=8107f192ffac406380b6d61d3d3dbf7d).  After the data are downloaded and extracted, the Excel file with the raw data is named "Peconics SCDHS WQ data - up to 2019 so far.xlsx", or something similar depending on when the data were downloaded.  The location of this file on your computer is passed to the import function. Below, a local file renamed as "currentdata.xlsx" that contains the water quality data is downloaded. 

```{r}
dat <- read_pepwq('../inst/extdata/currentdata.xlsx')
head(dat)
```

The raw data includes multiple fields, but only the chlorophyll and secchi data are retained for reporting.  Each station is grouped by major bay segment, defined as `r levels(dat$bay_segment)`. 

A quick view of the number of observations and length of record at each station shows that effort was not continuous. 

```{r, fig.height = 8, fig.width = 11}
toplo <- dat %>% 
  select(bay_segment, BayStation, yr, chla, sd) %>% 
  pivot_longer(c('chla', 'sd')) %>% 
  group_by(bay_segment, BayStation, yr, name) %>% 
  summarise(`Obs. (n)` = n())

p <- ggplot(toplo, aes(x = yr, y = BayStation, fill = `Obs. (n)`)) + 
  geom_tile()+ #colour = 'lightgrey') + 
  facet_grid(bay_segment ~ name, scales = 'free_y', space = 'free_y') + 
  theme_bw() + 
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor= element_blank(),
    strip.background = element_blank(), 
    axis.title.x = element_blank(), 
    legend.position = 'top', 
    axis.text.y = element_text(size = 7)
  ) + 
  scale_fill_viridis_c()

p
```

The function `anlz_pepdat()` summarizes the station data by bay segment.  The funciton returns annual means for chlorophyll and secchi depth and monthly means by year for chlorophyll and secchi depth. These summaries are then used to determine if bay segment targets for water quality are met using the anlz_attain() and `anlz_attainpep()` function.

Below shows how to use `anlz_pepdat()` to summarize the data by bay segment to estimate annual and monthly means for chlorophyll and secchi depth. The output is a two-element list for the annual (ann) and monthly (mos) means by segment.

```{r}
medpep <- anlz_medpep(dat)
medpep
```

This output can then be further analyzed with `anlz_attainpep()` to determine if the bay segment outcomes are met in each year. The results are used by the plotting functions described below. In short, the `chl_sd` column indicates the categorical outcome for chlorophyll and light attenuation for each segment. The outcomes are integer values from zero to three. The relative exceedances of water quality thresholds for each segment, both in duration and magnitude, are indicated by higher integer values.

```{r}
anlz_attainpep(medpep)
```

